\documentclass[10pt]{article}
\usepackage{array}
\usepackage[margin=2.5cm]{geometry}
\usepackage{longtable}
\usepackage{hyperref}
\usepackage{url}
\usepackage{enumitem}
\date{}

\newcolumntype{L}{>{\raggedleft}p{0.14\textwidth}}
\newcolumntype{R}{p{0.8\textwidth}}

%%%%%%%%%%%%%%%% Course Number/Title %%%%%%%%%%%%%%%%
\title{\vspace{-50pt}\bfseries STAT 111: Understanding Statistics \vspace{-50pt}}

\begin{document}
\maketitle
\setlength{\parindent}{0pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%% Logistics %%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\line(1,0){475}\\
%%%%%%%%%%%%%%%% Course Info %%%%%%%%%%%%%%%%
\begin{minipage}[t]{0.5\textwidth}
  \begin{flushleft}
    \begin{tabular}{l r}
      \textbf{Course:} & STAT 111\\
      \textbf{Section:} & 001\\
      \textbf{Semester:} & Summer 2015\\
      \textbf{Lecture Room:} & HOD-D 259\\
      \textbf{Lab Room:} & EIS G33\\
      \textbf{Meeting Time:} & MWTHF 12:00pm - 1:15pm\\
    \end{tabular}
  \end{flushleft}
\end{minipage}
%%%%%%%%%%%%%%%% Instructor Info %%%%%%%%%%%%%%%%
\begin{minipage}[t]{0.5\textwidth}
  \begin{flushright}
    \begin{tabular}{l r}
      \textbf{Instructor:} & Doug Raffle\\
      \textbf{Office:} & 423 Stansbury Hall\\
      \textbf{Office Hours:} & MTWH 1:30-3:00pm\\ & or by Appointment\\
      \textbf{Email:} &
        \href{mailto:dcraffle@mix.wvu.edu}{dcraffle@mix.wvu.edu}\\
      \textbf{Website:} & \url{stat.wvu.edu/~draffle/}\\
    \end{tabular}
  \end{flushright}
\end{minipage}\\
\vspace{10pt}
\line(1,0){475}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%% Course Background %%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\textbf{Course Description:} Introduction to basic concepts and ideas of statistics. Methodologies and case studies to prepare students to understand the use of statistics in the mass media and professional publications in their major field of study.\\

\textbf{Learning Outcomes:} At the end of this course, students wil be
able to
\begin{enumerate}[nolistsep]
  \item Use graphical and numerical summaries to describe data.
  \item Investigate relationships between variables.
  \item Compare and contrast different types of studies.
  \item Use probability to account for randomness and uncertainty.
  \item Perform and interpret statistical tests used to study
    populations.
  \item Compare groups using statistical tests.
\end{enumerate}

\vspace{10pt}

\textbf{Required Text}\\
\begin{tabular}{l p{10cm}}
  \textbf{Title:} & \emph{Intro Stats}\\
  \textbf{Author:} & Richard D. De Veaux, Paul F. Velleman, David E. Bock\\
  \textbf{Edition:} & 4th\\
  \textbf{Publisher:} & Pearson\\
  \textbf{Note:} & An electronic version of this text is included with
    MyStatLab.
\end{tabular}\\
\vspace{10pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%% Minimum Student Requirements %%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{Required Materials:} Students must have access to a copy of
the required text and a \href{mystatlab.com}{MyStatLab} account.
Access codes can be purchased from the bookstore or directly
from Pearson at \href{mystatlab.com}{MyStatLab.com}.  Students also
have the option to get started with free temporary access, but a key must be
purchased within 17 days.\\


\textbf{Required Technology:} Students must have a modern web browser
compatible with Pearson's \href{mystatlab.com}{MyStatLab}.  Use the
browser check tool to verify compatibility.\\

\textbf{Lab Sessions}
On Fridays, the class will meet in Eiesland G33 for a laboratory
session.  These labs will give students the chance to get hands-on
experience working with data.\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%% Grading %%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{Grading Policies:} Final letter grades are based on the total number
of points earned out of a possible 400 as follows:
\begin{center}\begin{tabular}{rl}
  A: & 360--400 points\\
  B: & 320--359 points\\
  C: & 280--319 points\\
  D: & 240--279 points\\
  F: & $<$ 239 points\\
\end{tabular}\end{center}

\textbf{Course Assessments:} Points earned towards the final total are
based on a final exam, homework assignments, quizzes, and lab assignments.
\begin{center}\begin{tabular}{l r r}
  & \multicolumn{2}{c}{\textbf{Points}} \\\hline
  \textbf{Assessment} & \textbf{Each} & \textbf{Total}\\\hline
  Assignments (10) & 10 & 100\\
  Quizzes (5) & 20 & 100\\
  Lab Assignments (5) & 20 & 100 \\
  Final Exam &  & 100 \\\hline
  \textbf{Grand Total} & & \textbf{400}\\
\end{tabular}\\\end{center}

\textbf{Assessment Details:}
\begin{enumerate}
  \item \textbf{Assignments} will require students to implement and
    discuss the techniques presented in this course.  All homework
    assignments are assigned through MyStatLab.  There will be a total
    of twelve homeworks assigned (two per week) and students will be
    graded on the best ten.
  \item \textbf{Quizzes:} At least six quizzes will be given throughout the
    course of the session (typically at the start of lab sessions).
    Students will be evaluated on the best five.
  \item \textbf{Labs Assignments:} Lab assignments will be require
    students to analyze real data sets and summarize the results of
    these analyses.  Lab 
    assignments need to be completed during the lab session to receive
    credit unless otherwise specified. At least six lab assignments will be
    given and students graded on the best five.
  \item \textbf{Final Exam:} At the end of the session a comprehensive
    final exam will be given to assess students' mastery of
    topics covered.
\end{enumerate}

\textbf{Course Schedule:} Assignments will become available in
MyStatLab as topics are covered.  The schedule on my website will also
be constantly updated.\\

\textbf{Attendance/Make-up Policy:} Attendance will not be taken, but poor
lecture attendance in this course nearly always results in low grades.
Make-ups will not be given for missed lab assignments or quizzes.  If
you need to miss a lab for any reason, the missed assignment and quiz
will count as your drops.\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%% Obligatory Disclaimers %%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{Academic Integrity:} The integrity of the classes offered by any academic institution
solidifies the foundation of its mission and cannot be sacrificed to
expediency, ignorance, or blatant fraud. Therefore, I will enforce
rigorous standards of academic integrity in all aspects and
assignments of this course. For the detailed policy of West Virginia
University regarding the definitions of acts considered to fall under
academic dishonesty and possible ensuing sanctions, please see the
Student Conduct Code at\\
\url{http://studentlife.wvu.edu/studentconductcode.html}. Should you
have any questions about possibly improper research citations or 
references, or any other activity that may be interpreted as an
attempt at academic dishonesty, please see me before the
assignment is due to discuss the matter.\\

\textbf{Inclusivity Statement}:
The West Virginia University community is committed to creating and
fostering a positive learning and working environment based on open
communication, mutual respect, and inclusion. If you are a person with
a disability and anticipate needing any type of accommodation in order
to participate in this class, please advise me and make appropriate
arrangements with the Office of Accessibility Services (293-6700). For
more information on West Virginia University’s Diversity, Equity, and
Inclusion initiatives, please see \url{http://diversity.wvu.edu}.\\

\textbf{Support Services}
Students can obtain academic and technical support at the following links:
\begin{itemize}
    \item WVU Student Support Services \url{http://sss.wvu.edu/}
    \item WVU Library \url{http://libraries.wvu.edu}
    \item Blackboard Accessibility
      \url{http://www.blackboard.com/Platforms/Learn/Resources/Accessibility.aspx }
    \item WVU Academic Catalog \url{http://catalog.wvu.edu/}
    \item WVU Accessibility Services \url{http://accessibilityservices.wvu.edu/}
    \item WVU Computer Security \url{http://it.wvu.edu/security}
\end{itemize}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%% Course Schedule %%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}