---
title: 'Lab 1: Motor Trend Cars'
output: html_document
---

## Overview
In this lab, we will examine the Motor Trend Cars data set.

The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

Variable | Description
---------|---------------
mpg |	Miles/(US) Gallon
cyl |	Number of Cylinders
disp |	Displacement (cu.in.)
hp |	Gross horsepower
wt |	Weight (lb/1000)
qsec |	1/4 mile time
vs |	V or Straight Configuration Engine
am |	Transmission Type (auto/manual)

## Load the Data in StatCrunch
A Tab Separated File (.tsv) containing the data can be found here:

[http://stat.wvu.edu/~draffle/111/week1/lab1/mtcars2.tsv](http://stat.wvu.edu/~draffle/111/week1/lab1/mtcars2.tsv)

1. In StatCrunch, use the menus `Data` $\to$ `Load` $\to$ `From File` $\to$ `On the web`.
2. Copy and Paste the address given above into `WWW Address`
4. Click `Load File` at the bottom of the page.

## Complete the following problems on a separate piece of paper.
Hand in your answers at the end of lab -- make sure to include you name and the lab number.  Instructions are included for making graphs and finding the necessary statistics for each problem.

Each question will be worth two points, and you can receive partial credit for incorrect answers if your process was correct.

Write your final answer as a sentence and include all steps you used to get there, otherwise you will receive partial credit.

### 1.) List each variable and its type (categorical/quantitative). 
```{r}
library(ggplot2)
mtcars2 <- read.table("mtcars2.tsv", header = T, sep = "\t")
data.frame(Variable = colnames(mtcars2)[-1], 
           Type = c("Numeric", "Categorical", "Numeric", "Numeric", "Numeric",
                    "Numeric", "Categorical", "Categorical"))
```

### 2.) Report the Frequency and Relative Frequency Tables of `cyl`.  Describe the distribution.

```{r}
counts <- summary(factor(mtcars2$cyl))
n <- nrow(mtcars2)
data.frame(Frequency = counts, 
           "Rel. Freq" = counts/n*100)
```

Most of the cars in the data set have eight cylinders, while six is the least common.

### 3.) Sketch a Bar Plot of `am`.  Which Transmission type is more common?

```{r}
ggplot(mtcars2, aes(x = am)) + geom_bar()
```

Manual cars are more common than automatics in this data set.

### 4.) Make a Histogram of `hp` (you do not need to sketch it).  Describe the shape of the distribution.

```{r}
ggplot(mtcars2, aes(x = hp)) + geom_histogram(binwidth = 50, color = "grey80")
```

The distribution of horsepower is right-skewed and unimodal with no outliers.

### 5.) Which Measure of Center is better for `hp`?  Report it.

```{r}
summary(mtcars2$hp)
```

Because the distribution is right-skewed, the median (123 hp) is a more accurate measure of center.

### 6.) Sketch a Side-by-Side Boxplot of `hp` grouped by `am`.  Describe the differences in the distributions.

```{r}
ggplot(mtcars2, aes(x = am, y = hp, fill = am)) + geom_boxplot()
```

Manuals have higher horsepower on average and more variation, while there are two automatics with unusually high horsepower.

### 7.) Sketch the QQ Plot of `mpg`.  Is the distribution approximately Normal?
```{r, echo = FALSE}
gg_qq <- function(X, x){
  y <- quantile(X$mpg, c(0.25, 0.75))
  x <- qnorm(c(0.25, 0.75))
  slope <- diff(y)/diff(x)
  int <- y[1L] - slope * x[1L]
  p <- ggplot(X, aes(sample=mpg)) +
        stat_qq(size = 2) +
        geom_abline(slope = slope, intercept = int, color="blue") +
        xlab("Normal Quantile") + ylab("Observed Quantile")
  return(p)
}
```

```{r message = FALSE, warning = FALSE}
library(car)
gg_qq(mtcars2, "mpg")
```

Because fuel efficiency generally follows the reference line, we can conclude that it is approximately normal.

### 8.) If a certain car had a fuel efficiency of 22 MPG, would it be more rare for an automatic or a manual car?  Show your work.

```{r}
by(mtcars2, mtcars2$am, function(tr) round(data.frame(mean = mean(tr$mpg), sd = sd(tr$mpg)), 2))
(z.auto <- (22 - 24.39)/6.17)
(z.manual <- (22 - 17.15)/3.83)
```

22 mpg would be 1.27 standard deviations above the mean for manuals, and 0.39 standard deviations below the mean for automatics.  This means that it would be more rare for a manual car to get 22 mpg.

### 9.) Assume that `MPG` $\sim N(\mu = 20, \sigma = 6)$.  What is the 90% percentile for the `MPG` of all cars?  
I.e., if $P(MPG \le x) = 0.9$, what is $x$?

```{r}
qnorm(.9, mean = 20, sd = 6)
```

The 90% of cars have fuel efficiency less than 27.7 mpg.

### 10.) If quarter mile times follow the Normal Model $N(\mu = 17.85, \sigma = 1.79)$, how quickly do the middle 75% of cars complete a quarter mile drag race?  
I.e., if $P(l \le QSEC \le u) = 0.75$, what are $l$ and $u$? 

```{r}
qnorm(c(.125, .875), mean = 17.85, sd = 1.79)
```

The middle 75% of cars finish the quarter mile in 15.8-19.9 seconds.












